Shrinking Characteristics of Precision Matrix Estimators: An Illustration via Regression

### 2019-04-20 01-intro.Rmd
intro, PME final

Inferential statistics consists of two primary components: estimation and prediction. Estimation refers to the process by which we infer values, properties and behavior about individual parameters in our model. Prediction, on the other hand, refers to the process by which we draw inference about an event that has not-yet occurred. This component is prevalent in industry where the demand to predict stock prices, to predict which movie individuals will prefer, and even new friend connections in a network has sparked huge investments in research and given rise to groups whose primary goal is pushing the bounds and developing new predictive models.

However, though most attention is directed towards prediction, these two components are not distinct. In fact, the success and reliability of many predictive models is contingent upon good, efficient parameter estimation. Many of these models require the estimation of a precision matrix - the inverse of the covariance matrix (frequently denotes as $\Omega$) - that establishes the interaction and covariance between random variables.

For this reason, in the last decade there has been an ever-expanding community devoted to precision matrix estimation and among this community is Professor Adam Rothman and Aaron Molstad, Ph.D. whose research will be a focal point of this manuscript. The two have published work on indirect multivariate linear regression (@molstad2016indirect) and classification with matrix-valued predictors (@molstad2018penalized) but the focus of this work is their 2017 paper titled *Shrinking Characteristics of Precision Matrix Estimators* (@molstad2017shrinking). In it, they outline a framework to estimate *characteristics* of precision matrices - a concept that exploits the fact that in many predictive models estimation of the precision matrix is only necessary through its product with another feature. They write in their manuscript that "to fit many predictive models, only a characteristic of the population precision matrix needs to be estimated. For example, in binary linear discriminant analysis, the population precision matrix is needed for prediction only through the product of the precision matrix and the difference between the two conditional distribution mean vectors." The purpose of the research detailed here began with the desire to expand on this concept and to explore avenues that were mentioned but were not further investigated.

One of the research directions mentioned in the original paper was the application of their framework to regression. Utilizing the fact that the population regression coefficient matrix $\beta \equiv \Omega_{x}\Sigma_{xy}$ (where $\Sigma_{xy}$ is the cross-covariance matrix between the predictors, $X$, and the responses, $Y$, and $\Omega_{x}$ is the precision matrix for $X$), their framework allows for the simultaneous estimation of $\beta$ and $\Omega_{x}$ with an embedded assumption potentially useful for superior prediction performance. In close communication and collaboration with Professor Rothman, we wanted to explore this direction further. However, in order to build upon their work and contribute new material, there were a number of concepts that needed to be learned along the way and this document will follow that journey.

We will begin chapter two with a brief introduction to precision matrix estimation and the gaussian log-likelihood function. This section will mention popular estimation methods and algorithms but most discussion will be directed towards the ADMM algorithm. This discussion will be useful as we begin detailing the shrinking characteristics of precision matrix estimators framework (which may be referred to as SCPME), the so-called augmented ADMM algorithm, and later the framework's application to regression. Lastly, the document will end with two brief tutorials for the R packages `ADMMsigma` and `SCPME`. These packages were developed by myself to aid in simulation experiments and make it easier to branch into related research directions. Both packages have since been published on CRAN.

@@

Most matrices will take the form of either $\Sigma$, the population covariance matrix, or $\Omega$, the population precision matrix. Note that the precision matrix is simply the inverse of the covariance matrix ($\Omega \equiv \Sigma^{-1}$) and a subscript may be added to each if more than two random variables are considered in a problem. A subscript star may also be added if the object is oracle - or known - a priori ($\Omega_{*}$). The oracle's estimator that minimizes/maximizes a pre-specified objective function will be denoted with a hat ($\hat{\Omega}$).

>>>>

Inferential statistics consists of two primary components: estimation and prediction. Estimation refers to the process by which we infer values, properties and behavior about individual parameters in our model. Prediction, on the other hand, refers to the process by which we draw inference about an event that has not-yet occurred. The latter component is abundant in industry where the demand to predict stock prices, movies individuals prefer, and even to predict new friend connections in a network has sparked huge investments in research and given rise to research groups whose primary goal is pushing the bounds and developing new predictive models.

However, though most attention is directed towards prediction, these two components are not distinct. In fact, the success and reliability of many predictive models is contingent upon good, efficient parameter estimation. Many of these models require the estimation of a precision matrix - the inverse of the covariance matrix (frequently denoted as $\Omega$) - that establishes the interaction and covariance between random variables. For this reason, the last decade has seen an ever-expanding community devoted to precision matrix estimation.

Among this community of researchers is Professor Adam Rothman and Aaron Molstad, Ph.D. whose research will be a focal point of this manuscript. The two have published work on indirect multivariate linear regression (@molstad2016indirect) and classification with matrix-valued predictors (@molstad2018penalized) but the focus of this manuscript is their 2017 paper titled *Shrinking Characteristics of Precision Matrix Estimators* (@molstad2017shrinking). In it, they outline a framework to estimate *characteristics* of precision matrices - a concept that exploits the fact that in many predictive models estimation of the precision matrix is only necessary through its product with another feature. They write in their manuscript that "to fit many predictive models, only a characteristic of the population precision matrix needs to be estimated... In binary linear discriminant analysis, the population precision matrix is needed for prediction only through the product of the precision matrix and the difference between the two conditional distribution mean vectors." The purpose of the research detailed here began with the desire to expand on this concept and to explore avenues that were mentioned but were not further investigated.

One of the research directions mentioned in the original paper was the application of their framework to regression. Utilizing the fact that the population regression coefficient matrix $\beta \equiv \Omega_{x}\Sigma_{xy}$ (where $\Sigma_{xy}$ is the cross-covariance matrix between the predictors, $X$, and the responses, $Y$, and $\Omega_{x}$ is the precision matrix for $X$), their framework allows for the simultaneous estimation of $\beta$ and $\Omega_{x}$ with an embedded assumption potentially useful for superior prediction performance. In close communication and collaboration with Professor Rothman, we wanted to explore this research direction further. However, in order to build upon their work and contribute new material, there were a number of concepts that needed to be learned along the way and this document will follow that journey.

We will begin chapter two with a brief introduction to precision matrix estimation and the gaussian log-likelihood function. This section will mention popular estimation methods and algorithms but most discussion will be directed towards the ADMM algorithm. Discussion of the ADMM algorithm will be useful as we begin detailing the shrinking characteristics of precision matrix estimators framework (which may be referred to as SCPME), the so-called augmented ADMM algorithm, and later the framework's application to regression. Lastly, the document will end with two brief tutorials for the R packages `ADMMsigma` and `SCPME`. These packages were developed by myself to aid in simulation experiments and make it easier to branch into related research directions. Both packages have since been published on CRAN.

@@

Most matrices will take the form of either $\Sigma$, the population covariance matrix, or $\Omega$, the population precision matrix. Note that the precision matrix is simply the inverse of the covariance matrix ($\Omega \equiv \Sigma^{-1}$) and a subscript may be added to each if more than two random variables are considered in a problem ($\Omega_{x}$). A subscript star may also be added if the object is oracle - or known - a priori ($\Omega_{*}$). The oracle's estimator that optimizes a pre-specified objective function will be denoted with a hat ($\hat{\Omega}$).

### 2019-04-20 02-PME.Rmd
intro, PME final

The estimator for $\mu$ that maximizes the log-likelihood is $\hat{\mu}^{MLE} = \bar{X} \equiv \sum_{i = 1}^{n}X_{i}/n$, so that the partially maximized gaussian log-likelihood for $\Omega$ is

@@

where $S = \sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n$ is the usual sample estimator for the covariance matrix, $\Sigma$. In addition to $\mu$, we could also derive the maximum likelihood estimator for $\Omega$. By setting the gradient of the partially maximized log-likelihood equal to zero, we can show that

\begin{equation}
\begin{split}
\hat{\Omega}^{MLE} &= \arg\max_{\Omega \in S_{+}^{p}}\left\{ \frac{n}{2}\log\left| \Omega \right| - \frac{n}{2}tr\left(S\Omega \right) \right\} \\
  &= \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| \right\} \\
  &= S^{-1}
\end{split}
(\#eq:omegamle)\notag
\end{equation}

so that the MLE for $\Omega$, when it exists, is $\hat{\Omega}^{MLE} = S^{-1} = \left[\sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n \right]^{-1}$. The reality, however, is that this object does *not* always exist. In settings where the number of observations is exceeded by the number of features in a sample, the sample covariance matrix is rank deficient and no longer invertible. For this reason, many papers in the last decade have proposed shrinkage estimators of the precision matrix similar to the shrinkage estimators in regression settings. That is, instead of minimizing solely the negative log-likelihood function, authors have proposed minimizing the gaussian log-likelihood *plus* a penalty term, $P$, which is often a function of the precision matrix.

@@

The penalties that have been proposed for precision matrix estimation are typically a variation of the ridge penalty $P\left(\Omega \right) = \lambda\|\Omega \|_{F}^{2}/2$ or the lasso penalty $P\left(\Omega \right) = \lambda\left\|\Omega\right\|_{1}$. The authors, @yuan2007model, initially proposed the lasso-penalized gaussian log-likelihood defined as

@@

so as not to penalize the diagonal elements of the precision matrix estimate. Other papers published on the lasso-penalized gaussian likelihood precision matrix estimator include @rothman2008sparse and @friedman2008sparse. In addition, many efficient algorithms have been proposed to solve for $\hat{\Omega}$ in this setting but the most popular method is the graphical lasso algorithm (often referred to as glasso) introduced by @friedman2008sparse. Their method utilizes an iterative block-wise coordinate descent algorithm that builds upon the coordinate descent algorithm used in lasso-penalized regression.

Non-lasso, non-convex penalties were considered in @lam2009sparsistency and @fan2009network and other papers considered penalizations like the Frobenius norm (@rothman2014existence; @witten2009covariance; @price2015ridge). In fact, the latter two papers show that the resulting minimizer can be solved in closed-form - which will be discussed later in this manuscript. However, the penalty explored in the remainder of this chapter is not a lasso penality nor a ridge penalty but, in fact, a convex combination of the two known as the *elastic-net* penalty:

@@

with tuning parameter $0 \leq \alpha \leq 1$. Clearly, when $\alpha = 0$ this penalty reduces to a ridge penalty and when $\alpha = 1$ it reduces to a lasso penalty. Originally proposed in @zou2005regularization in the context of regression, this penalty has since been popularized and is used in the penalized regression R package `glmnet`. This penalty allows for additional flexibility but, despite this, no published work (to our knowledge) has explored the elastic-net penalty in the context of precision matrix estimation. We will show how to solve the following optimization problem in the next section using the ADMM algorithm.

@@

In general, suppose that we want to solve an optimization problem of the form

@@

In this formulation, $y \in \mathbb{R}^{p}$ is called the lagrange multiplier and $\rho > 0$ is some scalar that acts as the step size for the algorithm. Note that any infimum under the augmented lagrangian is equivalent to the infimum of the traditional lagrangian since any feasible point $(x, z)$ must satisfy the constraint $\rho\left\| Ax + Bz - c \right\|_{2}^{2}/2 = 0$. @boyd2011distributed show that the infimum will be approached under the following repeated iterations:

@@

where the superscript, $k$, denotes the number of iterations. Conveniently, this general algorithm can be coerced into a format useful in precision matrix estimation. Suppose we let $f$ be equal to the non-penalized log-likelihood, $g$ equal to the elastic-net penalty, $P\left( \Omega \right)$, and we use the constraint that $\Omega$ must be equal to $Z$, then the augmented lagrangian in the context of precision matrix estimation is of the form

@@

where $\Lambda$ is taking the role of $y$ and the ADMM algorithm now consists of the repeated iterations

@@

It turns out that each step in this algorithm can be solved in closed-form. The full details of each can be found in the appendix \@ref(proofomegaalgo) but the following theorem provides the simplified steps in the algorithm.

@@

Define the soft-thresholding function as $\mbox{soft}(a, b) = \mbox{sign}(a)(\left| a \right| - b)_{+}$. Set $k = 0$ and initialize $Z^{0}, \Lambda^{0}$, and $\rho$. Repeat steps 1-3 until convergence:

@@

Another popular, alternative form of the ADMM algorithm can be used when scaling the dual variable ($\Lambda^{k}$). We derive this scaled-form in the context of precision matrix estimation here.

@@

Therefore, a scaled-form can now be written as

\begin{equation}
\begin{split}
  \Omega^{k + 1} &= \arg\min_{\Omega \in \mathbb{R}_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| + \frac{\rho}{2}\left\| \Omega - Z^{k} + U^{k} \right\|_{F}^{2} \right\} \\
  Z^{k + 1} &= \arg\min_{Z \in \mathbb{S}^{p}}\left\{ \lambda\left[ \frac{1 - \alpha}{2}\left\| Z \right\|_{F}^{2} + \alpha\left\| Z \right\|_{1} \right] + \frac{\rho}{2}\left\| \Omega^{k + 1} - Z + U^{k} \right\|_{F}^{2} \right\} \\
  U^{k + 1} &= U^{k} + \Omega^{k + 1} - Z^{k + 1}
(\#eq:ADMMscaled)
\end{split}
\end{equation}

@@

As a proof-of-concept that the elastic-net penalty in the context of precision matrix estimation can provide useful results and that the ADMM algorithm used in this process works, the section offers a short simulation. For the simulation, we generated data from multiple, unique precision matrices with various oracle structures. For each data-generating procedure, the algorithm was run with a 5-fold cross validation. After 20 replications, the cross validation errors were totalled and the optimal tuning parameters were selected (results are in the top half of the figures). These results were then compared with the Kullback Leibler (KL) losses between the estimated matrices and the oracle matrices (results are in the bottom half of the figures).

@@

The second figure shows the results when the data was generated from a multivariate normal distribution with mean equal to zero and a *dense* oracle precision matrix. Here, we randomly generated an orthogonal basis, set all eigen values equal to 1000, and then combined the matrices using QR decomposition. Interestingly, we find that the optimal $\alpha$ in this case is 0.6 which closely matches the optimal result based on the KL loss. This shows that there are cases where an elastic-net penalty can provide useful results and that using only a lasso penalty may unnecessarily restrict our penalized estimation.

>>>>

The estimator for $\mu$ that maximizes the log-likelihood is $\hat{\mu}^{mle} = \bar{X} \equiv \sum_{i = 1}^{n}X_{i}/n$, so that the partially maximized gaussian log-likelihood function for $\Omega$ is

@@

where $S = \sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n$ is the usual sample estimator for the population covariance matrix, $\Sigma$. In addition to $\mu$, one could also derive the maximum likelihood estimator for $\Omega$. By setting the gradient of the partially maximized log-likelihood equal to zero, one could show that

\begin{equation}
\begin{split}
\hat{\Omega}^{mle} &= \arg\max_{\Omega \in S_{+}^{p}}\left\{ \frac{n}{2}\log\left| \Omega \right| - \frac{n}{2}tr\left(S\Omega \right) \right\} \\
  &= \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| \right\} \\
  &= S^{-1}
\end{split}
(\#eq:omegamle)\notag
\end{equation}

so that the MLE for $\Omega$, when it exists, is $\hat{\Omega}^{mle} = S^{-1} = \left[\sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n \right]^{-1}$. The reality, however, is that this object does *not* always exist. In settings where the number of observations is exceeded by the number of features in a sample, the sample covariance matrix is rank deficient and no longer invertible. For this reason, many papers in the last decade have proposed *shrinkage estimators* of the population precision matrix similar to the shrinkage estimators in regression settings. That is, instead of minimizing solely the negative log-likelihood function, researchers have proposed minimizing the gaussian log-likelihood *plus* a penalty term, $P$, where $P$ is often a function of the precision matrix.

@@

The penalties that have been proposed for precision matrix estimation are typically a variation of the ridge penalty $P\left(\Omega \right) = \lambda\|\Omega \|_{F}^{2}/2$ or the lasso penalty $P\left(\Omega \right) = \lambda\left\|\Omega\right\|_{1}$ (here $\lambda$ is a tuning parameter). The authors, @yuan2007model, initially proposed the lasso-penalized gaussian log-likelihood defined as

@@

so as not to penalize the diagonal elements of the precision matrix estimate. Other papers published on the lasso-penalized gaussian likelihood precision matrix estimator include @rothman2008sparse and @friedman2008sparse. In addition, many efficient algorithms have been proposed to solve for $\hat{\Omega}$, however, the most popular method is the graphical lasso algorithm (glasso) introduced by @friedman2008sparse. Their method utilizes an iterative block-wise coordinate descent algorithm that builds upon the coordinate descent algorithm used in lasso-penalized regression.

Non-lasso, non-convex penalties were considered in @lam2009sparsistency and @fan2009network and other papers considered penalizations like the Frobenius norm (@rothman2014existence; @witten2009covariance; @price2015ridge). In fact, the latter two papers show that the resulting minimizer can be solved in closed-form - which will be discussed later in this manuscript. However, the penalty explored through the remainder of this chapter is not a lasso penality nor a ridge penalty but, in fact, a convex combination of the two known as the *elastic-net* penalty:

@@

with additional tuning parameter $0 \leq \alpha \leq 1$. Clearly, when $\alpha = 0$ this penalty reduces to a ridge penalty and when $\alpha = 1$ it reduces to a lasso penalty. Originally proposed in @zou2005regularization in the context of regression, this penalty has since been popularized and is used in the penalized regression R package `glmnet`. This penalty allows for additional flexibility but, despite this, no published work to our knowledge has explored the elastic-net penalty in the context of precision matrix estimation. We will show how to solve the following optimization problem in the next section using the ADMM algorithm.

@@

In general, the ADMM algorithm supposes that we want to solve an optimization problem of the form

@@

In this formulation, $y \in \mathbb{R}^{p}$ is called the lagrange multiplier and $\rho > 0$ is some scalar that acts as the step size for the algorithm. Note that any infimum under the augmented lagrangian is equivalent to the infimum of the traditional lagrangian since any feasible point $(x, z)$ must satisfy the constraint $\rho\left\| Ax + Bz - c \right\|_{2}^{2}/2 = 0$. @boyd2011distributed show that using the ADMM algorithm the infimum will be approached under the following repeated iterations:

@@

where the superscript, $k$, denotes the number of iterations. Conveniently, this general algorithm can be coerced into a format useful in precision matrix estimation. Suppose we let $f$ be equal to the non-penalized gaussian log-likelihood, $g$ equal to the elastic-net penalty, $P\left( \Omega \right)$, and we use the constraint that $\Omega \in \mathbb{S}_{+}^{p}$ must be equal to some matrix $Z \in \mathbb{R}^{p \times p}$, then the augmented lagrangian in the context of precision matrix estimation is of the form

@@

where $\Lambda$ takes the role of $y$ as the lagrange multiplier. The ADMM algorithm now consists of the following repeated iterations:

@@

Furthermore, it turns out that each step in this algorithm can be solved efficiently in closed-form. The full details of each can be found in the appendix \@ref(proofomegaalgo) but the following theorem provides the simplified steps in the algorithm.

@@

Define the soft-thresholding function as $\mbox{soft}(a, b) = \mbox{sign}(a)(\left| a \right| - b)_{+}$ and $S$ as the sample covariance matrix. Set $k = 0$ and initialize $Z^{0}, \Lambda^{0}$, and $\rho$. Repeat steps 1-3 until convergence:

@@

Another popular, alternative form of the ADMM algorithm can be used when scaling the dual variable ($\Lambda^{k}$) which we will briefly mention in the context of precision matrix estimation here.

@@

Therefore, a scaled-form ADMM algorithm can now be written as

\begin{equation}
\begin{split}
  \Omega^{k + 1} &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| + \frac{\rho}{2}\left\| \Omega - Z^{k} + U^{k} \right\|_{F}^{2} \right\} \\
  Z^{k + 1} &= \arg\min_{Z \in \mathbb{S}^{p}}\left\{ \lambda\left[ \frac{1 - \alpha}{2}\left\| Z \right\|_{F}^{2} + \alpha\left\| Z \right\|_{1} \right] + \frac{\rho}{2}\left\| \Omega^{k + 1} - Z + U^{k} \right\|_{F}^{2} \right\} \\
  U^{k + 1} &= U^{k} + \Omega^{k + 1} - Z^{k + 1}
(\#eq:ADMMscaled)
\end{split}
\end{equation}

@@

As a proof-of-concept that the elastic-net penalty in the context of precision matrix estimation can provide useful results and that the ADMM algorithm used in this process works, this section offers a short simulation. For the simulation, we generated data from multiple, unique precision matrices with various oracle structures. For each data-generating procedure, the algorithm was run with a 5-fold cross validation to tune parameters $\lambda$ and $\alpha$. After 20 replications, the cross validation errors were totalled and the optimal tuning parameters were selected (results are in the top half of the figures). These results were then compared with the Kullback Leibler (KL) losses between the estimated matrices and the oracle matrices (results are in the bottom half of the figures).

@@

The second figure shows the results when the data was generated from a multivariate normal distribution with mean equal to zero and a *dense* oracle precision matrix (non-sparse). Here, we randomly generated an orthogonal basis, set all eigen values equal to 1000, and then combined the matrices using QR decomposition. Interestingly, we find that the optimal $\alpha$ in this case is 0.6 which closely matches the optimal result based on the KL loss. This shows that there are cases where an elastic-net penalty can provide useful results and that using only a lasso penalty may unnecessarily restrict our penalized estimation.
